---
title: "Cape Cod Run Expectancy 288 2021-2023"
author: "Richard Legler"
date: "June 6, 2024"
output: 
  rmdformats::readthedown:
    number_sections: true
    highlight: tango
    df_print: paged
    center: true
---

```{r setup, include=FALSE}
# Set working directory
knitr::opts_knit$set(root.dir = 'C:/Users/richa/OneDrive/Documents/Hyannis Harbor Hawks/Cape Code Data')

# Clear environment
rm(list=ls())

# Import libraries
library(dplyr)
library(tidyr)
library(openxlsx)
```

# Import Data

```{r}
# Import
df_2011_2022 <- read.csv("CCBL full_pbp 2011-2022.csv")
df_2023 <- read.csv("CCBL full_pbp 2023.csv")

# Keep Necessary Columns
df_2011_2022 <- df_2011_2022[,c("Date", "GameID","Inning", "Top_Btm", "Balls", "Strikes", "Outs", "Runner1B", "Runner2B", "Runner3B","InningPACount", "PitchNumPA", "TeamRuns", "RunsScoredOnPlay")]
df_2023 <- df_2023[,c("GameID","Inning", "Top_Btm", "Balls", "Strikes", "Outs", "Runner1B", "Runner2B", "Runner3B","InningPACount", "PitchNumPA", "TeamRuns", "RunsScoredOnPlay")]
```

# Transform Data
## Transform Date to Year

```{r}
# Transform Date to Year
df_2011_2022 <- df_2011_2022 %>%
  mutate(Year = format(as.Date(Date, format = "%m/%d/%Y"), "%Y")) %>%
  select(-Date) %>%
  rename(Date = Year)

# Add Year Column to 2023 Data
df_2023[,c("Date")] <- "2023"
```

## Remove Rows that add a Pinch Runner

Only occurs in 2023 PBP Data

```{r}
selected_row <- c()

for (i in 1:(nrow(df_2023))) {
  if (is.na(df_2023$Outs[i])){
    selected_row <- c(selected_row, i)
  }
}

df_2023 <- df_2023[-selected_row,]
```

## Merge Data for 2021 through 2023

```{r}
# Merge data frames
df <- rbind(df_2011_2022, df_2023)

# Transform Year to integer
df$Date <- as.integer(df$Date)

# Keep data from 2021 through 2023
df <- df %>%
  filter(Date >= 2021)
```


## Add Column with Score After the Pitch

```{r}
# Add column
df[,c("TeamRunsAfter")] <- df$TeamRuns + df$RunsScoredOnPlay
```

## Convert Runners on Base Data
```{r}
# First Base
df$Runner1B <- ifelse(is.na(df$Runner1B),0,1)

# Second Base
df$Runner2B <- ifelse(is.na(df$Runner2B),0,1)

# Third Base
df$Runner3B <- ifelse(is.na(df$Runner3B),0,1)
```

## Make Sure 3 outs in Inning

```{r}
df <- df %>%
  arrange(Date, GameID, Top_Btm, Inning, InningPACount, PitchNumPA) %>%
  mutate(OutsOnPlay = (lead(Outs) - Outs) %% 3) %>%
  group_by(GameID, Inning, Top_Btm) %>%
  mutate(OutsInInning = sum(OutsOnPlay)) %>%
  filter(OutsInInning == 3) %>%
  ungroup()
```


# Establish Game States
Game State format: Balls/Strikes/Outs/Runner on 1B/Runner on 2B/Runner on 3B
Examples:
  0-0 count, 0 outs and 0 baserunners -> 000000,
  2-1 count, 1 out and runner on 2nd -> 211010,
  1-2 count, 2 outs and bases loaded -> 122111

```{r}
# Add column for Game States
df[,c("GameState")] <- paste(df$Balls, df$Strikes, df$Outs, df$Runner1B, df$Runner2B, df$Runner3B, sep = "")
```

## Remove Duplicated Game States in a Plate Appearance

Occurs when a pitch is fouled off with 2 strikes, since the game state does not change

```{r}
selected_row <- c()

for (i in 1:(nrow(df) - 1)) {
  if (!is.na(df$GameState[i]) && !is.na(df$GameState[i + 1]) &&
      df$GameState[i] == df$GameState[i + 1] &&
      df$InningPACount[i] == df$InningPACount[i + 1]) {
    selected_row <- c(selected_row, i)
  }
}

df_clean <- df[-selected_row,]
```


# Runs Scored Rest of Inning
## Add Column for Runs Scored Rest of Inning

```{r}
df_clean[,c("RunsRestInning")] <- NA
```

## Seperate Data into Half Innings

```{r}
split_df <- split(df_clean, f=list(df_clean$GameID, df_clean$Inning, df_clean$Top_Btm, df_clean$Date))
```

## Function for runs scored in rest of inning

```{r}
runs_rest_inning <- function(current_score, score_end_inning){
  return(score_end_inning - current_score)
}
```

## Loop to assign runs scored

```{r}
# Loop through list of data frames
for (i in seq_along(split_df)) {
  split_data <- split_df[[i]]
  
  # Check if split_df is not empty
  if (nrow(split_data) > 0) {
  
    # Number of rows in subset
    n_rows <- nrow(split_data)
    
    # Score at end inning of the subset
    score_end_inning <- tail(split_data$TeamRunsAfter, 1)
    
    # Loop through each row in subset
    for (j in 1:n_rows) {
      # Retrieve current score
      current_score <- split_data$TeamRuns[j]
      
      # Assign runs of rest inning
      split_data$RunsRestInning[j] <- runs_rest_inning(current_score, score_end_inning)
    }
    
    # Update original dataframe with modified split_df
    df_clean[df_clean$GameID %in% split_data$GameID & 
         df_clean$Inning %in% split_data$Inning &
         df_clean$Top_Btm %in% split_data$Top_Btm, ] <- split_data
  }
}
```

# Run Expectancy

```{r}
game_state_summary <- df_clean %>%
  group_by(GameState) %>%
  summarize(
    count = n(), # Count of occurences
    sum_runs_rest_inning = sum(RunsRestInning, na.rm = TRUE), # Sum of runs_rest_inning
    expected_runs <- sum_runs_rest_inning / count # Expected runs in each game state
  ) %>%
  ungroup()

# Rename column
game_state_summary <- game_state_summary %>%
  rename(expected_runs = `expected_runs <- sum_runs_rest_inning/count`)
```

## Split out game states
For readability

```{r}
game_state_summary <- game_state_summary %>%
  mutate(balls = substr(GameState, 1, 1),
         strikes = substr(GameState, 2, 2),
         outs = substr(GameState, 3, 3),
         first = substr(GameState, 4, 4),
         second = substr(GameState, 5, 5),
         third = substr(GameState, 6, 6))
```

## Run Expectancy Matrices by Count

```{r}
# List to Store Run Expectancy Matrices
RE_matrices <- list()

# List of all possible counts
ball_count <- list(0, 1, 2, 3)
strike_count <- list(0, 1, 2)

# Convert game_state_summary data types
game_state_summary$balls <- as.integer(game_state_summary$balls)
game_state_summary$strikes <- as.integer(game_state_summary$strikes)
```


```{r}
# Loop Through Counts
for (i in 1:length(ball_count)) {
  for (j in 1:length(strike_count)){
    # Filtered game_state_summary
    temp <- game_state_summary %>%
      filter(balls == ball_count[i], strikes == strike_count[j])
    
    # Initialize and populate RE_matrix
    rows <- c("000", "100", "010", "001", "110", "101", "011", "111")
    columns <- c("0 outs", "1 out", "2 outs")
    RE_matrix <- matrix(NA, nrow = length(rows), ncol = length(columns), dimnames = list(rows, columns))
    
    match_indices_row <- match(substr(temp$GameState, 4, 6), rows)
    match_indices_col <- match(substr(temp$GameState, 3, 3), substr(columns, 1, 1))
    RE_matrix[cbind(match_indices_row, match_indices_col)] <- round(temp$expected_runs, 3)
    
    # Save matrix to list
    RE_matrices[[paste0("RE_matrix_", ball_count[i], "-", strike_count[j],"_count")]] <- RE_matrix
  }
}
```


# Export
## Game State Summary
```{r}
# Save the game_state_summary to a xlsx file
write.xlsx(game_state_summary, file = "C:/Users/richa/OneDrive/Documents/Hyannis Harbor Hawks/Run Expectancy & Steal Model/Output/RE_summary_by_count_2021x2023.xlsx", sheetName = "GameStateSummary", rowNames = FALSE, overwrite = TRUE)
write.xlsx(game_state_summary, file = "C:/Users/richa/OneDrive/Documents/Hyannis Harbor Hawks/Cape Code Data/RE_summary_by_count_2021x2023.xlsx", sheetName = "GameStateSummary", rowNames = FALSE, overwrite = TRUE)
```


## Matrix by Count

```{r}
# Create a new Excel workbook
wb <- createWorkbook()

# Loop through each count in RE_matrices
for (i in 1:length(ball_count)) {
  for (j in 1:length(strike_count)){
  
    # RE_matrix for current count
    RE_matrix <- RE_matrices[[paste0("RE_matrix_", ball_count[i], "-", strike_count[j],"_count")]]
    
    # Create a new sheet with the name "RE YYYY"
    sheetName <- paste(ball_count[i], strike_count[j], sep = "-")
    addWorksheet(wb, sheetName)
    
    # Add an initial column for the runners situation
    runners_situation <- c("000", "100", "010", "001", "110", "101", "011", "111")
    writeData(wb, sheet = sheetName, x = cbind(Runners_Situation = runners_situation, RE_matrix))
  }
}

# Save the workbook to a xlsx file
saveWorkbook(wb, "C:/Users/richa/OneDrive/Documents/Hyannis Harbor Hawks/Run Expectancy & Steal Model/Output/RE_matrices_by_count_2021x2023.xlsx", overwrite = TRUE)
saveWorkbook(wb, "C:/Users/richa/OneDrive/Documents/Hyannis Harbor Hawks/Cape Code Data/RE_matrices_by_count_2021x2023.xlsx", overwrite = TRUE)
```

