---
title: "Cape Cod Run Expectancy & Steal Model 2011-2022"
author: "Richard Legler"
date: "May 13, 2024"
output: 
  rmdformats::readthedown:
    number_sections: true
    highlight: tango
    df_print: paged
    center: true
---

```{r setup, include=FALSE}
# Set working directory
knitr::opts_knit$set(root.dir = 'C:/Users/richa/OneDrive/Documents/Hyannis Harbor Hawks/Cape Code Data')

# Clear environment
rm(list=ls())

# Import libraries
library(dplyr)
library(tidyr)
library(openxlsx)
```

# Import Data

```{r}
# Import
df <- read.csv("CCBL full_pbp 2011-2022.csv")

# Keep Necessary Columns
df <- df[,c("Date", "GameID","Inning", "Top_Btm", "Outs", "Runner1B", "Runner2B", "Runner3B","InningPACount", "PitchNumPA", "TeamRuns", "RunsScoredOnPlay")]
```

# Transform Data
## Transform Date to Year

```{r}
df <- df %>%
  mutate(Year = format(as.Date(Date, format = "%m/%d/%Y"), "%Y")) %>%
  select(-Date) %>%
  rename(Date = Year)
```

## Add Column with Score After the Pitch

```{r}
# Add column
df[,c("TeamRunsAfter")] <- df$TeamRuns + df$RunsScoredOnPlay
```

## Convert Runners on Base Data
```{r}
# First Base
df$Runner1B <- ifelse(is.na(df$Runner1B),0,1)

# Second Base
df$Runner2B <- ifelse(is.na(df$Runner2B),0,1)

# Third Base
df$Runner3B <- ifelse(is.na(df$Runner3B),0,1)
```

## Make Sure 3 outs in Inning

```{r}
df <- df %>%
  arrange(Date, GameID, Top_Btm, Inning, InningPACount, PitchNumPA) %>%
  mutate(OutsOnPlay = (lead(Outs) - Outs) %% 3) %>%
  group_by(GameID, Inning, Top_Btm) %>%
  mutate(OutsInInning = sum(OutsOnPlay)) %>%
  filter(OutsInInning == 3) %>%
  ungroup()
```


# Establish Game States
Game State format: Outs/Runner on 1B/Runner on 2B/Runner on 3B
Examples:
  0 outs and 0 baserunners -> 0000,
  1 out and runner on 2nd -> 1010,
  2 outs and bases loaded -> 2111

```{r}
# Add column for Game States
df[,c("GameState")] <- paste(df$Outs, df$Runner1B, df$Runner2B, df$Runner3B, sep = "")
```

## Remove Duplicated Game States in a Plate Appearance

```{r}
selected_row <- c()

for (i in 1:(nrow(df) - 1)) {
  if (!is.na(df$GameState[i]) && !is.na(df$GameState[i + 1]) &&
      df$GameState[i] == df$GameState[i + 1] &&
      df$InningPACount[i] == df$InningPACount[i + 1]) {
    selected_row <- c(selected_row, i)
  }
}

df_clean <- df[-selected_row,]
```


# Runs Scored Rest of Inning
## Add Column for Runs Scored Rest of Inning

```{r}
df_clean[,c("RunsRestInning")] <- NA
```

## Seperate Data into Half Innings

```{r}
split_df <- split(df_clean, f=list(df_clean$GameID, df_clean$Inning, df_clean$Top_Btm, df_clean$Date))
```

## Function for runs scored in rest of inning

```{r}
runs_rest_inning <- function(current_score, score_end_inning){
  return(score_end_inning - current_score)
}
```

## Loop to assign runs scored

```{r}
# Loop through list of data frames
for (i in seq_along(split_df)) {
  split_data <- split_df[[i]]
  
  # Check if split_df is not empty
  if (nrow(split_data) > 0) {
  
    # Number of rows in subset
    n_rows <- nrow(split_data)
    
    # Score at end inning of the subset
    score_end_inning <- tail(split_data$TeamRunsAfter, 1)
    
    # Loop through each row in subset
    for (j in 1:n_rows) {
      # Retrieve current score
      current_score <- split_data$TeamRuns[j]
      
      # Assign runs of rest inning
      split_data$RunsRestInning[j] <- runs_rest_inning(current_score, score_end_inning)
    }
    
    # Update original dataframe with modified split_df
    df_clean[df_clean$GameID %in% split_data$GameID & 
         df_clean$Inning %in% split_data$Inning &
         df_clean$Top_Btm %in% split_data$Top_Btm, ] <- split_data
  }
}
```

# Seperate Data into Respective Years

```{r}
for (year in 2011:2022) {
  # Filter dataframe for the current year
  filtered_df <- df_clean %>%
    filter(Date == as.character(year))
  
  # Assign the filtered dataframe to a new dataframe with year in name
  assign(paste0("df_", year), filtered_df)
}
```


# Run Expectancy


```{r}
# List to Store Run Expectancy Matrices
RE_matrices <- list()

# Loop Through Years
for (year in 2011:2022) {
  # Get dataframe for the current year
  df_year <- get(paste0("df_", year))
  
  # Check if dataframe is empty
  if (nrow(df_year) > 0) {
    # Perform RE_matrix calculations
    game_state_summary <- df_year %>%
      group_by(GameState) %>%
      summarize(
        count = n(),
        sum_runs_rest_inning = sum(RunsRestInning, na.rm = TRUE),
        expected_runs = sum_runs_rest_inning / count
      ) %>%
      ungroup()
    
    # Initialize and populate RE_matrix
    rows <- c("000", "100", "010", "001", "110", "101", "011", "111")
    columns <- c("0 outs", "1 out", "2 outs")
    RE_matrix <- matrix(NA, nrow = length(rows), ncol = length(columns), dimnames = list(rows, columns))
    
    match_indices_row <- match(substr(game_state_summary$GameState, 2, 4), rows)
    match_indices_col <- match(substr(game_state_summary$GameState, 1, 1), substr(columns, 1, 1))
    RE_matrix[cbind(match_indices_row, match_indices_col)] <- round(game_state_summary$expected_runs, 3)
    
    # Save RE_matrix to list with appropriate name
    RE_matrices[[paste0("RE_matrix_", year)]] <- RE_matrix
  }
}

print(RE_matrices)
```


# Steal Breakeven %
## Functions

```{r}
# Expected Runs Starting Game State
exp_run_start <- function(outs, runners, RE_matrix){
  # Row
  if(runners == "000"){
    row_index <- 1
  }
  else if(runners == "100"){
    row_index <- 2
  }
  else if(runners == "010"){
    row_index <- 3
  }
  else if(runners == "001"){
    row_index <- 4
  }
  else if(runners == "110"){
    row_index <- 5
  }
  else if(runners == "101"){
    row_index <- 6
  }
  else if(runners == "011"){
    row_index <- 7
  }
  else if(runners == "111"){
    row_index <- 8
  }
  
  # Columns
  col_index <- outs + 1
    
  # Corresponding expected runs
  result <- RE_matrix[row_index, col_index]
  return(as.numeric(result[1]))
}
```


```{r}
# Expected Runs Successful Game State
exp_run_success <- function(outs, runners, sb, RE_matrix){
  # Row
  if(runners == "000"){
    row_index <- 1
  }
  else if(runners == "100"){
    row_index <- 2
  }
  else if(runners == "010"){
    row_index <- 3
  }
  else if(runners == "001"){
    row_index <- 4
  }
  else if(runners == "110"){
    row_index <- 5
  }
  else if(runners == "101"){
    row_index <- 6
  }
  else if(runners == "011"){
    row_index <- 7
  }
  else if(runners == "111"){
    row_index <- 8
  }
  
  # Columns
  col_index <- outs + 1
    
  # Corresponding expected runs
  result <- RE_matrix[row_index, col_index]
  
  # Add a run if Home is stolen
  if(sb == "H"){
    return(as.numeric(result[1]+1))
  }
  else{
    return(as.numeric(result[1]))
  }
}
```


```{r}
# Expected Runs Failed Game State
exp_run_fail <- function(outs, runners, RE_matrix){
  # Add an out for being caught stealing
  new_outs <- outs + 1
  
  # End inning if 3rd out
  if(new_outs == 3){
    return(0)
  }
  else{
  # Row
  if(runners == "000"){
    row_index <- 1
  }
  else if(runners == "100"){
    row_index <- 2
  }
  else if(runners == "010"){
    row_index <- 3
  }
  else if(runners == "001"){
    row_index <- 4
  }
  else if(runners == "110"){
    row_index <- 5
  }
  else if(runners == "101"){
    row_index <- 6
  }
  else if(runners == "011"){
    row_index <- 7
  }
  else if(runners == "111"){
    row_index <- 8
  }
  
  # Corresponding expected runs
  result <- RE_matrix[row_index, new_outs+1]
  return(as.numeric(result[1]))
  }
}
```


```{r}
# Steal Success Rate (%) to Breakeven
breakeven_func <- function(start, success, fail){
  rate <- 1/((success - start)/(start - fail)+1)
  return(round(rate,3))
}
```

# Loop Through Years

```{r}
# Rows and columns of base data frame
runners <- rep(c("100", "010", "001", "110", "110", "101", "101", "011", "111"), each = 3)
outs <- rep(c(0, 1, 2), times = length(runners) / 3)
er_start <- rep(c(NA), times = length(runners))
base_to_steal <- rep(c("2", "3", "H","3", "23", "2", "H", "H", "H"), each = 3)
success <- rep(c("010", "001", "000", "101", "011", "011", "100", "010", "110"), each = 3)
er_success <- rep(c(NA), times = length(runners))
fail <- rep(c("000", "000", "000", "100", "010", "001","100","010","110"), each = 3)
er_fail <- rep(c(NA), times = length(runners))
breakeven <- rep(c(NA), times = length(runners))
```


```{r}
for (year in 2011:2022) {
  # Check if the RE_matrix for the current year exists in RE_matrices
  re_matrix_name <- paste0("RE_matrix_", year)
  if (re_matrix_name %in% names(RE_matrices)) {
    # Create a new data frame for the current year
    assign(paste0("steal_", year), data.frame(Runners = runners, Outs = outs, ER_Start = er_start, Stolen_Base = base_to_steal, Success = success, ER_Success = er_success, Fail = fail, ER_Fail = er_fail, Breakeven = breakeven))
    
    # Update matrix to be used
    RE_matrix_year <- RE_matrices[[re_matrix_name]]
    
    # Get the data frame for the current year
    steal_year <- get(paste0("steal_", year))

    for (i in 1:nrow(steal_year)) {
      # Update Expected Runs in Starting Game State
      steal_year$ER_Start[i] <- exp_run_start(steal_year$Outs[i], steal_year$Runners[i], RE_matrix_year)
      # Update Expected Runs in Success Game State
      steal_year$ER_Success[i] <- exp_run_success(steal_year$Outs[i], steal_year$Success[i], steal_year$Stolen_Base[i], RE_matrix_year)
      # Update Expected Runs in Fail Game State
      steal_year$ER_Fail[i] <- exp_run_fail(steal_year$Outs[i], steal_year$Fail[i], RE_matrix_year)
      # Update Breakeven Steal %
      steal_year$Breakeven[i] <- breakeven_func(steal_year$ER_Start[i], steal_year$ER_Success[i], steal_year$ER_Fail[i])
    }

    # Update the data frame in the environment
    assign(paste0("steal_", year), steal_year)
  } else {
    print(paste(re_matrix_name, "not found in RE_matrices. Skipping year."))
  }
}
```

## Summary

```{r}
for (year in 2011:2022){
  if(year == 2020){
   print("Season cancelled due to COVID-19.") 
  }
  else{
    print(paste0("Breakeven steal % for ", year, "."))
    print(get(paste0("steal_", year)))
  }
}
```

# Export
## Run Expectancy Matrices by Year

```{r}
# Create a new Excel workbook
wb <- createWorkbook()

# Loop through each year in RE_matrices
for (year in 2011:2022) {
  # RE_matrix for current year
  RE_matrix <- RE_matrices[[paste0("RE_matrix_", year)]]
  
  # Create a new sheet with the name "RE YYYY"
  sheetName <- paste("RE", year, sep = " ")
  addWorksheet(wb, sheetName)
  
  # Add an initial column for the runners situation
  runners_situation <- c("000", "100", "010", "001", "110", "101", "011", "111")
  writeData(wb, sheet = sheetName, x = cbind(Runners_Situation = runners_situation, RE_matrix))
}

# Save the workbook to a xlsx file
saveWorkbook(wb, "C:/Users/richa/OneDrive/Documents/Hyannis Harbor Hawks/Cape Code Data/RE_matrices_by_year.xlsx", overwrite = TRUE)
saveWorkbook(wb, "C:/Users/richa/OneDrive/Documents/Hyannis Harbor Hawks/Run Expectancy & Steal Model/Output/RE_matrices_by_year.xlsx", overwrite = TRUE)
```

## Steal Breakeven by Year

```{r}
# Create a new Excel workbook
wb_steal <- createWorkbook()

# Loop through each year in steal_YYYY dataframes
for (year in 2011:2022) {
  # Check if steal_YYYY dataframe exists
  if (exists(paste0("steal_", year))) {
    # Retrieve the steal_YYYY dataframe for the current year
    current_df <- get(paste0("steal_", year))

    # Create a new sheet with the name "steal YYYY"
    sheetName <- paste("steal", year, sep = " ")
    addWorksheet(wb_steal, sheetName)

    # Write the current dataframe to the sheet
    writeData(wb_steal, sheet = sheetName, x = current_df)
  }
}

# Save the workbook to an Excel file
saveWorkbook(wb_steal, "C:/Users/richa/OneDrive/Documents/Hyannis Harbor Hawks/Cape Code Data/steal_by_year.xlsx", overwrite = TRUE)
saveWorkbook(wb_steal, "C:/Users/richa/OneDrive/Documents/Hyannis Harbor Hawks/Run Expectancy & Steal Model/Output/steal_by_year.xlsx", overwrite = TRUE)
```

