---
title: "Cape Cod Run Probability 2023"
author: "Richard Legler"
date: "May 22, 2024"
output: 
  rmdformats::readthedown:
    number_sections: true
    highlight: tango
    df_print: paged
    center: true
---

```{r setup, include=FALSE}
# Set working directory
knitr::opts_knit$set(root.dir = 'C:/Users/richa/OneDrive/Documents/Hyannis Harbor Hawks/Cape Code Data')

# Clear environment
rm(list=ls())

# Import libraries
library(dplyr)
library(tidyr)
library(openxlsx)
```

# Import Data

```{r}
# Import
df <- read.csv("CCBL full_pbp 2023.csv")

# Keep Necessary Columns
df <- df[,c("GameID","Inning", "Top_Btm", "Outs", "Runner1B", "Runner2B", "Runner3B","InningPACount", "PitchNumPA", "TeamRuns", "RunsScoredOnPlay")]
```

# Transform Data
## Remove Rows that add a Pinch Runner

```{r}
selected_row <- c()

for (i in 1:(nrow(df))) {
  if (is.na(df$Outs[i])){
    selected_row <- c(selected_row, i)
  }
}

df <- df[-selected_row,]
```


## Add Column with Score After the Pitch

```{r}
# Add column
df[,c("TeamRunsAfter")] <- df$TeamRuns + df$RunsScoredOnPlay
```

## Convert Runners on Base Data
```{r}
# First Base
df$Runner1B <- ifelse(is.na(df$Runner1B),0,1)

# Second Base
df$Runner2B <- ifelse(is.na(df$Runner2B),0,1)

# Third Base
df$Runner3B <- ifelse(is.na(df$Runner3B),0,1)
```

## Make Sure 3 outs in Inning

```{r}
df <- df %>%
  arrange(GameID, Top_Btm, Inning, InningPACount, PitchNumPA) %>%
  mutate(OutsOnPlay = (lead(Outs) - Outs) %% 3) %>%
  group_by(GameID, Inning, Top_Btm) %>%
  mutate(OutsInInning = sum(OutsOnPlay)) %>%
  filter(OutsInInning == 3) %>%
  ungroup()
```

# Establish Game States
Game State format: Outs/Runner on 1B/Runner on 2B/Runner on 3B
Examples:
  0 outs and 0 baserunners -> 0000,
  1 out and runner on 2nd -> 1010,
  2 outs and bases loaded -> 2111

```{r}
# Add column for Game States
df[,c("GameState")] <- paste(df$Outs, df$Runner1B, df$Runner2B, df$Runner3B, sep = "")
```

## Remove Duplicated Game States in a Plate Appearance

```{r}
selected_row <- c()

for (i in 1:(nrow(df) - 1)) {
  if (!is.na(df$GameState[i]) && !is.na(df$GameState[i + 1]) &&
      df$GameState[i] == df$GameState[i + 1] &&
      df$InningPACount[i] == df$InningPACount[i + 1]) {
    selected_row <- c(selected_row, i)
  }
}

df_clean <- df[-selected_row,]
```


# Are Runs Scored in the Rest of the Inning?
## Add Column for Runs Scored Rest of Inning

```{r}
df_clean[,c("RunsRestInning")] <- NA
```

## Seperate Data into Half Innings

```{r}
split_df <- split(df_clean, f=list(df_clean$GameID, df_clean$Inning, df_clean$Top_Btm))
```

## Function for runs scored in rest of inning

```{r}
runs_rest_inning <- function(current_score, score_end_inning){
  if(current_score == score_end_inning){
    return(FALSE)
  }
  else{
   return(TRUE) 
  }
}
```

## Loop to assign runs scored

```{r}
# Loop through list of data frames
for (i in seq_along(split_df)) {
  split_data <- split_df[[i]]
  
  # Check if split_df is not empty
  if (nrow(split_data) > 0) {
  
    # Number of rows in subset
    n_rows <- nrow(split_data)
    
    # Score at end inning of the subset
    score_end_inning <- tail(split_data$TeamRunsAfter, 1)
    
    # Loop through each row in subset
    for (j in 1:n_rows) {
      # Retrieve current score
      current_score <- split_data$TeamRuns[j]
      
      # Assign runs of rest inning
      split_data$RunsRestInning[j] <- runs_rest_inning(current_score, score_end_inning)
    }
    
    # Update original dataframe with modified split_df
    df_clean[df_clean$GameID %in% split_data$GameID & 
         df_clean$Inning %in% split_data$Inning &
         df_clean$Top_Btm %in% split_data$Top_Btm, ] <- split_data
  }
}
```

# Run Probability

```{r}
game_state_summary <- df_clean %>%
  group_by(GameState) %>%
  summarize(
    count = n(), # Count of occurences
    more_runs_in_inning = sum(RunsRestInning, na.rm = TRUE), # Count of runs_rest_inning
    run_probabilty = more_runs_in_inning / count # Expected runs in each game state
  ) %>%
  ungroup()
```

## Run Probability Chart

```{r}
# Rows and columns
rows <- c("000", "100", "010", "001", "110", "101", "011", "111")
columns <- c("0 outs", "1 out", "2 outs")

# Initialize matrix
RP_matrix <- matrix(NA, nrow = length(rows), ncol = length(columns), dimnames = list(rows, columns))

# Match rows and columns
match_indices_row <- match(substr(game_state_summary$GameState, 2, 4), rows)
match_indices_col <- match(substr(game_state_summary$GameState, 1, 1), substr(columns, 1, 1))

# Assign values
RP_matrix[cbind(match_indices_row, match_indices_col)] <- paste0(round(game_state_summary$run_probabilty * 100, 1), "%")
print(noquote(RP_matrix))
```

# Export
## Run Probability Matrix by Year

```{r}
# Create a new Excel workbook
wb <- createWorkbook()

# Create a new sheet with the name "RP 2023"
sheetName <- "RP 2023"
addWorksheet(wb, sheetName)
  
# Add an initial column for the runners situation
runners_situation <- c("000", "100", "010", "001", "110", "101", "011", "111")
writeData(wb, sheet = sheetName, x = cbind(Runners_Situation = runners_situation, RP_matrix))

# Save the workbook to a xlsx file
saveWorkbook(wb, "C:/Users/richa/OneDrive/Documents/Hyannis Harbor Hawks/Run Probability/Output/RP_matrix_2023.xlsx", overwrite = TRUE)
saveWorkbook(wb, "C:/Users/richa/OneDrive/Documents/Hyannis Harbor Hawks/Cape Code Data/RP_matrix_2023.xlsx", overwrite = TRUE)
```

